### 2018-03-19 Paper/jaccardbaetica.tex
update introduction

Material culture allows us to understand a part of mechanism of production... (aquí creo que me repito como siempre)

The economy of Roman Empire have been an object of study in the last centuries. (cambiar)

This paper aims to highlight the production dynamics in relation to a specific area within Roman Empire. We want to detect the pattern of olive oil production that link amphora workshops and amphoric stamps.
We focus here on exploring the economic relation between stamps and amphora production centres.

Roman provinces such as Baetica became important production and distribution centers during the Roman Empire. However, it remains under debate how this province was organized and whether it is possible to identify patterns in the olive oil market.

@@

All the amphora stamps belong to Dressel 20 types \citep{dressel,martin-kilcher_romischen_1994}. Dressel 20 is commonly associated with transportation of Baetican olive oil through the provinces during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps and \textit{tituli picti} and inked in graffities with different information but there is not a general consensus about the meaning of them \citep{rodriguez_baetican_1998}. Stamps are the most studied in this type of amphorae. There is evidence that stamps were used for almost three centuries. (economía oleica betica). Frequently, stamps were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}

The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria (CITAR). Stamps was mostly formed for a code of three letters. There letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

The meaning of the amphora stamps is still under debate. Some authors suggest that they were identified as the land-owners of the olive groves \citep{rodriguez_economioleicola_1977}. Other authors propose that stamp could be the owners of the making-amphorae workshop (CITAR) or even a production counting system \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps defined somehow the system of working in the workshops. 
%habría que citar a Dressel
%olive producing

>>>>

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. They allows us to highlight a part of mechanism of production and distribution of goods between different communities.

%Roman trade has been an object of study in the last centuries. 
%hablar de que diferentes modelos se han utilizado como el de brumel, el ikasen de las redes, xavi pero qeu lo vamos a probar a microescala para ver si rula o no.

%This paper explore--- resumir en dos líneas el objeto de estudio y definir que ha sido usado para ello
This paper aims to highlight the production dynamics in relation to a specific area within Roman Empire. We want to detect the pattern of olive oil production that link amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production centres. An ecology approach has been used to analyse the dispersion of stamps between amphora workshops.

% hipotesis

% Debate

%más específico--- olive oil

Roman province such as Baetica (currently Andalusia, south Spain) became important production centers of olive oil in the Roman Age. Olive oil was an important product during the Roman Empire used in almost all the aspect of daily life as cooking, hygiene or lightning \citep{mattingly_d.j._oil_1988}.

The good condition of the Baetican lands stimulated by the high demand of Roman provinces allowed to develop a massive infrastructure of olive oil production. This product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army (citar Carreras o Blazquez illo).

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries (cronologiaaa CITAR). As a results, hundred of amphora workshops were created to support this high demand. These workshops were located along the river Guadalquivir and its tributary Genil supplying from the riverine connectivity to the Mediterranean and Atlantic (CITAR GARCIA VARGAS)

The majority of olive oil amphorae produced in this province and shipped thought the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps and \textit{tituli picti} and inked in graffities with different information but there is not a general consensus about the meaning of them \citep{rodriguez_baetican_1998}. Stamps are the most studied in this type of amphorae. There is evidence that stamps were used for almost three centuries. (economía oleica betica). Frequently, stamps were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria (CITAR). Stamps was mostly formed for a code of three letters. There letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. 
The meaning of the amphora stamps is still under debate. Some authors suggest that they were identified as the land-owners of the olive groves \citep{rodriguez_economioleicola_1977}. Other authors propose that stamp could be the owners of the making-amphorae workshop (CITAR) or even a production counting system \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps defined somehow the system of working in the workshops.

%Hyphotesis
However, it remains under debate how this province was organized and whether it is possible to identify patterns in the olive oil trade.

@@
